Concurrent Programming in Harmony

### 2021-12-25 book/paper.tex
Fixed some stuff

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queueconc}
uses the <{alloc}> module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using <{malloc()}> and <{free()}>.
<{malloc(v}>) returns a new memory location initialized to <{v}>,
which should be released with <{free()}> when it is no longer in use.
The queue maintains a <{head}> pointer to the first element in the list
and a <{tail}> pointer to the last element in the list.
The <{head}> pointer is <{None}> if and only if the queue is empty.
(<{None}> is a special address value that is not the address of any
memory location.)

@@

\section*{Exercises}
\begin{problems}
% \item Assume that you have only binary semaphores.
% Implement counting semaphores using the split binary semaphores technique.
\item Several of the calls to <{release_one()}> in
\autoref{fig:RWsplitsema} can be replaced by simply releasing <{mutex}>.
Which ones?
\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
\item \label{ex:bbsbs} Implement a solution to the producer/consumer problem
using split binary semaphores.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to <{M}> threads to acquire it at the same time.\footnote{A bound lock is a restricted version of a \emph{counting} semaphore.}
A bound lock
with \texttt{M = 1} is an ordinary lock.
You should define a constant <{M}> and two methods:
<{acquire_bound_lock()}> and <{release_bound_lock()}>.
(Bound locks are useful for situations where too many threads working
at the same time might exhaust some resource such as a cache.)
\item Write a test program for your bound lock
that checks that no more than <{M}> threads can acquire the
bound lock.
\item Write a test program for bound locks
that checks that up to <{M}> threads
can acquire the bound lock at the same time.
% \item Implement a \emph{bounded stack}.  Method <{push(item}>)
% should block if the stack is full.  Method \texttt{pop()} should block
% until the stack is non-empty and then return an item.
% Also implement test programs for your solution.
\item \label{ex:gpu} Implement a thread-safe \emph{GPU allocator} by modifying
\autoref{fig:gpu}.
There are <{N}> GPUs identified by the numbers
1 through <{N}>.  Method <{gpuAlloc()}> returns the identifier
of an available GPU, blocking if there is currently no GPU available.
Method <{gpuRelease(gpu}>) releases the given GPU.  It never needs
to block.
\item With reader/writer locks,
concurrency can be improved if a thread \emph{downgrades} its write lock
to a read lock when its done writing but not done reading.  Add
a <{downgrade}> method to the code in \autoref{fig:RWsplitsema}.
(Similarly, you may want to try to implement an \emph{upgrade} of a
read lock to a write lock.  Why is this problematic?)
% How easy or hard would it be to do the same for the code in
% \autoref{fig:rw2lock}?
% Note, you cannot simply release the
% write lock and then obtain the read lock, because the thread needs
% to remain in the critical section.
\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
On a one-lane bridge,
cars can only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.
Write methods \texttt{OLBlock()} to create a new ``one lane bridge'' lock,
<{nb_enter()}> that a car must invoke before going northbound on
the bridge and <{nb_leave()}> that the car must invoke after leaving
the bridge.  Similarly write <{sb_enter()}> and <{sb_leave()}>
for southbound cars.
\item Extend the solution to \autoref{ex:onelane} by implementing the
requirement that at most $n$ cars are allowed on the bridge.  Add $n$
as an argument to \texttt{OLBlock}.

@@

The <{synch}> library supports a synchronized message queue,
similar to the \texttt{Queue} object in Python.
Its interface is as follows:
\begin{itemize}
\item \texttt{Queue()} returns a new message queue;
\item <{put(q}>, <{item}>) adds <{item}> to the queue pointed to by <{q}>;
\item <{get(q}>) waits for and returns an item on the queue pointed to by <{q}>.
\end{itemize}

@@

<{setintlevel(il}>) sets the interrupt level to <{il}> and returns
the prior interrupt level.  Returning the old level is handy when writing interrupt-safe
methods that can be called from ordinary code as well as from an interrupt handler.
\autoref{fig:trap5} shows how one might write a interrupt-safe method
to increment the counter.

@@

As an example, we will revisit the producer/consumer problem.
The code in \autoref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  <{items}> is an unbounded array with each
entry initialized to <{None}>.  <{back}> is an index into the
array and points to the next slot where a new value is inserted.
The code uses two atomic operations:
\begin{itemize}
\item <{inc(p}>): atomically increments \texttt{!}<{p}> and returns
the old value;
\item <{exch(p}>): sets \texttt{!}<{p}> to <{None}> and returns
the old value.
\end{itemize}

Method <{produce(item}>) uses <{inc}>(?<{back}>)
to allocate
the next available slot in the <{items}> array.
It stores the item as a singleton tuple.
Method <{consume()}> repeatedly scans the array, up to the
<{back}> index, trying to find an item to return.
To check an entry, it uses <{exch()}>
to atomically remove an item from a slot if there is one.
This way, if two or more threads attempt to extract an item from
the same slot, at most one will succeed.

@@

Method <{app_send(m}>) retransmits <{m}> until
an acknowledgment is received.
Method <{app_recv(reliable}>) returns the next successfully received
message.
\autoref{fig:abptest} shows how the methods may be used to send and receive
a stream of \texttt{NMSGS} messages reliably.
It has to be bounded, because model checking requires a finite model.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
<{tas(lk)}> & test-and-set on <{!lk}> \\
\hline
<{cas}(ptr, old, new)}> & compare-and-swap on <{!ptr}> \\
\hline
<{BinSem(v)}> & return a binary semaphore initialized to <{v}> \\
\hline
<{Lock()}> & return a binary semaphore initialized to <{False}> \\
\hline
<{acquire(bs)}> & acquire binary semaphore \texttt{!}<{bs}> \\
\hline
<{release(bs)}> & release binary semaphore \texttt{!}<{bs}> \\
\hline
<{Condition()}> & return a condition variable \\
\hline
<{wait(c}>, <{lk}>) & wait on condition variable \texttt{!}<{c}>
and lock <{lk}> \\ \hline
<{notify(c)}> & notify a thread waiting on condition variable \texttt{!}<{c}> \\
\hline
<{notifyAll(c)}> & notify all threads waiting on condition variable \texttt{!}<{c}> \\
\hline
<{Semaphore(cnt)}> & return a counting semaphore initialized to <{cnt}> \\
\hline
<{P(sema)}> & procure \texttt{!}<{sema}>  \\
\hline
<{V(sema)}> & vacate \texttt{!}<{sema}>  \\
\hline
\texttt{Queue}() & return a synchronized queue object \\
\hline
<{get(q}>) & return next element of <{q}>, blocking if empty \\
\hline
<{put(q, item)}> & add <{item}> to <{a}> \\
\hline
\end{tabular}

>>>>

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queueconc}
uses the <{alloc}> module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using <{malloc()}> and <{free()}>.
<{malloc(v)}> returns a new memory location initialized to <{v}>,
which should be released with <{free()}> when it is no longer in use.
The queue maintains a <{head}> pointer to the first element in the list
and a <{tail}> pointer to the last element in the list.
The <{head}> pointer is <{None}> if and only if the queue is empty.
(<{None}> is a special address value that is not the address of any
memory location.)

@@

\section*{Exercises}
\begin{problems}
% \item Assume that you have only binary semaphores.
% Implement counting semaphores using the split binary semaphores technique.
\item Several of the calls to <{release_one()}> in
\autoref{fig:RWsplitsema} can be replaced by simply releasing <{mutex}>.
Which ones?
\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
\item \label{ex:bbsbs} Implement a solution to the producer/consumer problem
using split binary semaphores.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to <{M}> threads to acquire it at the same time.\footnote{A bound lock is a restricted version of a \emph{counting} semaphore.}
A bound lock
with \texttt{M = 1} is an ordinary lock.
You should define a constant <{M}> and two methods:
<{acquire_bound_lock()}> and <{release_bound_lock()}>.
(Bound locks are useful for situations where too many threads working
at the same time might exhaust some resource such as a cache.)
\item Write a test program for your bound lock
that checks that no more than <{M}> threads can acquire the
bound lock.
\item Write a test program for bound locks
that checks that up to <{M}> threads
can acquire the bound lock at the same time.
% \item Implement a \emph{bounded stack}.  Method <{push(item)}>
% should block if the stack is full.  Method \texttt{pop()} should block
% until the stack is non-empty and then return an item.
% Also implement test programs for your solution.
\item \label{ex:gpu} Implement a thread-safe \emph{GPU allocator} by modifying
\autoref{fig:gpu}.
There are <{N}> GPUs identified by the numbers
1 through <{N}>.  Method <{gpuAlloc()}> returns the identifier
of an available GPU, blocking if there is currently no GPU available.
Method <{gpuRelease(gpu)}> releases the given GPU.  It never needs
to block.
\item With reader/writer locks,
concurrency can be improved if a thread \emph{downgrades} its write lock
to a read lock when its done writing but not done reading.  Add
a <{downgrade}> method to the code in \autoref{fig:RWsplitsema}.
(Similarly, you may want to try to implement an \emph{upgrade} of a
read lock to a write lock.  Why is this problematic?)
% How easy or hard would it be to do the same for the code in
% \autoref{fig:rw2lock}?
% Note, you cannot simply release the
% write lock and then obtain the read lock, because the thread needs
% to remain in the critical section.
\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
On a one-lane bridge,
cars can only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.
Write methods \texttt{OLBlock()} to create a new ``one lane bridge'' lock,
<{nb_enter()}> that a car must invoke before going northbound on
the bridge and <{nb_leave()}> that the car must invoke after leaving
the bridge.  Similarly write <{sb_enter()}> and <{sb_leave()}>
for southbound cars.
\item Extend the solution to \autoref{ex:onelane} by implementing the
requirement that at most $n$ cars are allowed on the bridge.  Add $n$
as an argument to \texttt{OLBlock}.

@@

The <{synch}> library supports a synchronized message queue,
similar to the \texttt{Queue} object in Python.
Its interface is as follows:
\begin{itemize}
\item \texttt{Queue()} returns a new message queue;
\item <{put(q, item)}> adds <{item}> to the queue pointed to by <{q}>;
\item <{get(q)}> waits for and returns an item on the queue pointed to by <{q}>.
\end{itemize}

@@

<{setintlevel(il)}> sets the interrupt level to <{il}> and returns
the prior interrupt level.  Returning the old level is handy when writing interrupt-safe
methods that can be called from ordinary code as well as from an interrupt handler.
\autoref{fig:trap5} shows how one might write a interrupt-safe method
to increment the counter.

@@

As an example, we will revisit the producer/consumer problem.
The code in \autoref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  <{items}> is an unbounded array with each
entry initialized to <{None}>.  <{back}> is an index into the
array and points to the next slot where a new value is inserted.
The code uses two atomic operations:
\begin{itemize}
\item <{inc(p)}>: atomically increments \texttt{!}<{p}> and returns
the old value;
\item <{exch(p)}>: sets \texttt{!}<{p}> to <{None}> and returns
the old value.
\end{itemize}

Method <{produce(item)}> uses <{inc(?back)}> to allocate
the next available slot in the <{items}> array.
It stores the item as a singleton tuple.
Method <{consume()}> repeatedly scans the array, up to the
<{back}> index, trying to find an item to return.
To check an entry, it uses <{exch()}>
to atomically remove an item from a slot if there is one.
This way, if two or more threads attempt to extract an item from
the same slot, at most one will succeed.

@@

Method <{app_send(m)}> retransmits <{m}> until
an acknowledgment is received.
Method <{app_recv(reliable)}> returns the next successfully received
message.
\autoref{fig:abptest} shows how the methods may be used to send and receive
a stream of \texttt{NMSGS} messages reliably.
It has to be bounded, because model checking requires a finite model.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
<{tas(lk)}> & test-and-set on <{!lk}> \\
\hline
<{cas}(ptr, old, new)}> & compare-and-swap on <{!ptr}> \\
\hline
<{BinSem(v)}> & return a binary semaphore initialized to <{v}> \\
\hline
<{Lock()}> & return a binary semaphore initialized to <{False}> \\
\hline
<{acquire(bs)}> & acquire binary semaphore \texttt{!}<{bs}> \\
\hline
<{release(bs)}> & release binary semaphore \texttt{!}<{bs}> \\
\hline
<{Condition()}> & return a condition variable \\
\hline
<{wait(c, lk)}> & wait on condition variable \texttt{!}<{c}>
and lock <{lk}> \\ \hline
<{notify(c)}> & notify a thread waiting on condition variable \texttt{!}<{c}> \\
\hline
<{notifyAll(c)}> & notify all threads waiting on condition variable \texttt{!}<{c}> \\
\hline
<{Semaphore(cnt)}> & return a counting semaphore initialized to <{cnt}> \\
\hline
<{P(sema)}> & procure \texttt{!}<{sema}>  \\
\hline
<{V(sema)}> & vacate \texttt{!}<{sema}>  \\
\hline
\texttt{Queue}() & return a synchronized queue object \\
\hline
<{get(q)}> & return next element of <{q}>, blocking if empty \\
\hline
<{put(q, item)}> & add <{item}> to <{a}> \\
\hline
\end{tabular}
